---
title: "R Mini Tutorial: Getting started and data types"
author: "luiscartor"
date: "May 29, 2019"
output: github_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Getting started

In Unix or Mac, execute R on the command line, or click on the R icon on the desktop for Windows. 

To finish an R session, type `q()` at the command prompt. You will be asked to save the workspace when finishing the session. It is a good practice to NOT to save the session, but if you say yes, the objects on your workspace will be saved as a .RData file. If saved, these workspace will be loaded when you open a new session. 

## 1.1 Basic calculations 

Use the console to execute R commands, such as basic calculations. After hitting enter, the console will show the result.

```{r}
  3+2
```

But the result of the calculation can be also stored to a variable:

```{r}
  a <- 3+2
```

Variable names in R must begin with a letter, followed by alphanumeric characters. Periods or underscores are often used to break up long variables (e.g. `long_variable`). Use variable names that help you identify the variable, but without being very long (e.g. herons_popmean). Avoid `c`, `q`, `t`, `C`, `D`, `F`, `I`, `T`, and things like `mean`, `var`, `sd`, as they are built-in R functions and they will cause errors. 

Now '5' is stored in the variable a. To print the value of a variable, just type the variable name:

```{r}
  a
```

Some arithmetic operators:

|Operator | Description |
|--- | --- |
|+| 	addition|
|-| 	subtraction|
|*| 	multiplication|
|/| 	division|
|^ or ** 	|exponentiation|

Some comparison and logical operators:

|Operator |	Description|
|--- | --- |
|> |	greater than|
|>= |	greater than or equal to|
|== |	exactly equal to|
|!= |	not equal to| 
|! |	logical NOT|
|& |	logical AND|


Now we can start creating variables and making arithmetic and logical operations. 

```{r}
x <- 5
y <- 2
z1 <- x*y
z2 <- x/y
z3 <- x^y
z1; z2; z3
```


Note that we can use a semicolon to put two or more commands in one line. In the last line of code, we use the semicolon to print the values of the three z variables.

We can combine several operations:

```{r}
A <- 5; B <- 3
C <- (A+B)/(A+5*B) 
C
```

And use the logical operators:

```{r}
A == B
```
```{r}
A > B
```
```{r}
C != B
```


We can use a wide rage of built-in mathematical functions in R

|R command |	function|
|--- | --- |
|abs() |	absolute value, $|
|cos(), sin(), tan() |	cosine, sine, tangent|
|acos(), asin(), atan() |	arc-cosine, arc-sine, arc-tangent|
|exp(x) |	exponential function|
|log() |	natural (base-e) logarithm|
|log10() |	base-10 logarithm|
|sqrt() |	square root, √x|

For example, try this calculations:

```{r}
A <- 5; B <- 3
C <- (A+2*sqrt(B))/(A+5*sqrt(B)) 
D <- log(A+B)/cos(C)
C;D
```


As usual, parentheses alter the order of the operations. Try:

```{r}
C <- A+2*sqrt(B)/A+5*sqrt(B)
C
```

and

```{r}
C <- A + 2*(sqrt(B)/A + 5) *sqrt(B)
C
```

<br>

> ### Excercise: Solve the following equations
>
> 1. $$1+0.3+0.7^2$$
> 2. For a = 0.5 and b = 1: $$sin(a)cos(b)-cos(a)$$ 
> 3. $$e^{0.5^2/4}$$
> 4. For a = 10 and b = 2: $$ \log(a)-3 * \log10(b) $$

<br>

## 1.2 Basic Data Types

### 1.2.1 Vectors

The simplest data type in R is the vector, which can be: numerical, character, logical, factor, or list based.   

```{r}
  a <- c(1,2,3,7.3,-2,-0.1) # numeric vector
  b <- c("one","two","three","a","b","c") # character vector
  c <- c(TRUE,TRUE,TRUE,FALSE,TRUE,FALSE) #logical vector
  a;b;c
```

We can access certain elements of the vector by using subscripts:

```{r}
a[5]
b[3]
a[c(1,3,5)]
```


Arithmetic expressions can be used with numerical vectors:

```{r}
  a <- a+1; a
  a <- sqrt(a); a
```

Also expressions involving multiple vectors:

```{r}
  a <- c(1,2,3,7.3,-2,-0.1)
  b <- c(1,0,1,0,2,2.5)
  c <- 2*a + b + sqrt(c) + 8
  c
```

What happens when the vectors have different lengths? The elements of the shorter vector are re-utilized for the elements of the long vector:

```{r}
  a <- c(10,20,30,40,50,60)
  b <- c(1,2,3)
  a + b
```

Some basic functions can be used to create vectors. One of this functions is `seq`, with syntax: `a <- seq(from, to, by)`:

```{r}
  a <- seq(1,9,2)
  a
```

Another useful function to create vectors is `rep` (`b <- rep(values,length)`):

```{r}
  b <- rep(3,5)
  b
```

### 1.2.2 Matrices

Now we are going to take a look how to create and handle matrices in R. All columns in a matrix must have the same data mode (numeric, character, logical, etc.) and the same length. A matrix is created using the following syntax: `m <- matrix(vector, nrow=r, ncol=c, byrow=FALSE, dimnames=list(rnames, cnames))`. See the example:

```{r}
  m <- matrix(1:20, nrow=5, ncol=4)
  m
```

Elements of the matrix can be accessed using subscripts:

```{r}
  m[,3] # 3th column of matrix
  m[4,] # 4rd row of matrix
  m[2:4,1:3] # rows 2,3,4 of columns 1,2,3 
```


A matrix can be also created by providing a vector and the row and column names:

```{r}
  v <- c(1,12,24,36)
  rnames <- c("R1", "R2") # character vector with row names
  cnames <- c("C1", "C2") # character vector with column names
  m <- matrix(v, nrow=2, ncol=2, byrow=TRUE,
  dimnames=list(rnames, cnames)) 
  m
```

An array is another type of matrix, but it can have more than two dimensions:
```{r}
v1 <- c(5,9,3)
v2 <- c(10,11,12,13,14,15)

arr <- array(c(v1,v2),dim = c(3,3,2))
arr
```


There are two very useful functions to deal with matrices, and that help us combine vectors of same size intro matrices, or combine matrices with vectors: `cbind` and `rbind`:

```{r}
  m1 <- cbind(1:3,4:6,5:7) 
  m1
```
```{r}
  m2 <- rbind(1:3,4:6)
  m2
```

<br>

> ### Excercises
> 1. Try to combine matrices m1 and m2 into one single matrix. What are the dimensions of the new matrix?
> 2. Subset the two first rows of m1 and combine the resulting matrix with m2. What are the dimensions of the new matrix?

<br>

### 1.2.2 Data frames

A data frame is similar to a matrix, but different columns can have different data modes (numeric, character, factor, etc.).

```{r}
  a <- c(1,2,3,7.3,-2,-0.1) # numeric vector
  b <- c("one","two","three","a","b","c") # character vector
  c <- c(TRUE,TRUE,TRUE,FALSE,TRUE,FALSE) #logical vector
  df <- data.frame(a,b,c)
  names(df) <- c("numbers","strings","passed")
  df
```

We can use different methods to identify elements of the data frame:

```{r}
df[1:2] # columns 2 and 3 of data frame
df[c("numbers","passed")] # columns numbers and passed 
df$strings # variable strings in the data frame 
```

### 1.2.3 Lists

A list can contain anything in each element: vectors, matrices, other lists and other objects:

```{r}
L <- list(A=a,B=b,C=c)
L
```

To access an element of the list, we need to use `[[ ]]` by number, or `$` to access an element by name:

```{r}
L[[1]] # Access elements A of the list by number
L[c("B","C")] # Access elements B and C of the list (access by element name)
```

### 1.2.4 Factors

A factor is a variable that can take a finite number of distinct levels. So it is a **nominal** variable. We can build a factor by applying a `factor` function to a vector of any class:

```{r}
v1 <- rep(c(1,2,3),each=3)
factor(v1)
```

| Useful functions | Description |
| --- | --- |
|length(object) | number of elements or components|
|str(object)    | structure of an object|
|class(object)  | class or type of an object|
|names(object)  | names|
|c(object,object,...)    | combine objects into a vector|
|cbind(object, object, ...)| combine objects as columns|
|rbind(object, object, ...)| combine objects as rows|
|object    | prints the object|
|ls()    | list current objects|
|rm(object) | delete an object|
|newobject <- edit(object)| edit copy and save as newobject|
|fix(object)   |   edit in place |

<br>

> ### Challenges
> 1. Try to access only the string "three" from the list L created previously.
> 2. Check the class type of each element of list L.
